---
title: "EDA Football Data UK"
output:
  html_document:
    df_print: paged
---


```{r setup}
library(tidyverse)
library(tidyr)

# knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_knit$set(root.dir = '') 
getwd()
```


```{r functions}

# functions ---------------------------------------------------------------
lm_eqn <- function(df){
  # fit regression
  model <- lm(final_tt ~ prob + I(prob^2), df)
  
  # get r2
  eq <- substitute(
    italic(r)^2 ~ "=" ~ r2,
    list(r2 = format(summary(model)$r.squared, digits = 3))
  )
  
  return(as.character(as.expression(eq)))
}


```


```{r readata, warning=FALSE}
# set working dir (must change)
# setwd(dir = "C:/Users/Ryo/Documents/Estudios/ITAM/Tesis/sportsAnalytics/tesis/Repository")

# folder 
folder_files <- "Data/football_uk/final_dbb/"

# read tables
df_r <- read_csv(file = paste0(folder_files, "footballuk_raw.csv"))
df_odds <- read_csv(file = paste0(folder_files, "stats_odds.csv"))
df_updateodds.pinnacle <- read_csv(file = paste0(folder_files, "odds_pinnacle.csv"))
df_varodds <- read_csv(file = paste0(folder_files, "variance_odds_deciles.csv"))
df_tt <- read_csv(file = paste0(folder_files, "track_take_books.csv"))
```

# General stats

```{r gen_stats}
# numebr of games per season
df_r %>% count(season)

# teams appareance by season
df_r %>% 
  count(hometeam, season) %>% 
  pivot_wider(
    names_from = season,
    names_glue = "season_{season}",
    values_from = n,
    values_fill = 0,
    values_fn = function(x) if_else(!is.na(x), 1, 0)
  ) %>% 
  mutate(
    number_seasons = rowSums(across(where(is.numeric)))
  )

# referee's appareance
df_r %>% 
  count(referee, season) %>% 
  arrange(referee, season, desc(n)) %>% 
  ungroup() %>% 
  group_by(season) %>% 
  arrange(season) %>% 
  pivot_wider(
    names_from = season,
    names_glue = "season_{season}",
    values_from = n,
    values_fill = 0,
    values_fn = function(x) ifelse(!is.na(x), x, 0)
  ) %>% 
  mutate(
    number_games = rowSums(across(where(is.numeric))),
    proportion_appearance = number_games / sum(number_games)
  ) %>% 
  arrange(desc(number_games))
```



# Odds

Look dataframe
```{r look_dfodds}
df_odds %>% head()
```


## Distribution Pinnacle Sports (Closing odds)

```{r plot_oddsdist}
# odds across seasons
df_r %>% 
  mutate(
    final_tt = 1/psch + 1/pscd + 1/psca - 1,
    
    psch = psch * (final_tt + 1),
    pscd = pscd * (final_tt + 1),
    psca = psca * (final_tt + 1)
  ) %>% 
  select(
    season, hometeam, awayteam, psch, pscd, psca 
  ) %>% 
  pivot_longer(
    cols = psch:psca,
    names_to = "type_odd",
    values_to = "odd"
  ) %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE)
  ) %>% 
  ggplot(aes(season, odd, color = type_odd)) +
  geom_boxplot(position = "dodge") +
  scale_y_log10() +
  ggtitle("Distribución de momios por temporadas") +
  xlab("Temporada") + ylab("Momio") +
  # guides(color = guide_legend(title = "Tipo de Momio")) +
  # scale_color_discrete(labels = c("Visitante", "Casa")) +
  theme_bw()




```

Look if odds are statistically the same across seasons. The rows are in the order:

- Home
- Draw
- Away

```{r odds_kruskal}
# get dataframe by type of odd
df_aux <- df_r %>% 
  mutate(
    final_tt = 1/psch + 1/pscd + 1/psca - 1,
    season = factor(season, levels = 13:20, ordered = TRUE),
    psch = psch * (final_tt + 1),
    pscd = pscd * (final_tt + 1),
    psca = psca * (final_tt + 1)
  ) %>% 
  select(
    season, hometeam, awayteam, psch, pscd, psca
  ) %>% 
  pivot_longer(
    cols = psch:psca,
    names_to = "type_odd",
    values_to = "odd"
  )

# test kruskal wallis each type of odd 
df_aux %>% 
  split(.$type_odd) %>% 
  map(
    ~ kruskal.test(
      odd ~ season,
      data = .
    )
  ) %>% 
  map_dfr(~ broom::tidy(.))




```

Look odds only by type of outcome of match. We saw that thre isn't evidence to think they are different along seasons at a significance of 1%.

```{r odds_outcome}
# by type off odd
df_r %>% 
  mutate(
    final_tt = 1/psch + 1/pscd + 1/psca - 1,
    
    psch = psch * (final_tt + 1),
    pscd = pscd * (final_tt + 1),
    psca = psca * (final_tt + 1)
  ) %>% 
  select(
    season, hometeam, awayteam, psch, pscd, psca
  ) %>% 
  pivot_longer(
    cols = psch:psca,
    names_to = "type_odd",
    values_to = "odd"
  ) %>% 
  ggplot(aes(type_odd, odd)) +
  geom_violin(fill = "gray50", width = 1.5) +
  geom_boxplot(width = 0.2, size = 1.2) +
  scale_y_continuous(trans = "log2", breaks = 2^(0:5), name = "Momio") +
  scale_x_discrete(labels = c("Visitante", "Empate", "Casa"), name = "") +
  ggtitle("Distribución de los momios", subtitle = "Momios finales corregidos de Pinnacle Sports") +
  theme_bw() +
  coord_flip() 


gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20, angle = 90),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25)
  ) 

# save
ggsave('momios_hda.png', plot = gph, device = png(), path = "Figures/Football_UK", width = 17.5, height = 10)
  
```




## Variablity

Look tables
```{r odds_var}
# generate table of corrected odds and probas
df_change_odds <- df_r %>% 
  mutate(
    match_name = paste(hometeam, awayteam, sep = "-"),
    
    season = factor(season, levels = 13:20, ordered = TRUE),
    
    apriori_tt = 1/psh + 1/psd + 1/psa - 1,
    posterior_tt = 1/psch + 1/pscd + 1/psca - 1,
    
    psh = psh * (apriori_tt + 1),
    psd = psd * (apriori_tt + 1),
    psa = psa * (apriori_tt + 1),
    
    psch = psch * (posterior_tt + 1),
    pscd = pscd * (posterior_tt + 1),
    psca = psca * (posterior_tt + 1),
    
    diff_home = psch/psh - 1,
    diff_draw = pscd/psd - 1,
    diff_away = psca/psa - 1
  )


# look
df_change_odds %>% head()
df_updateodds.pinnacle %>% head()
```


Look how log corrected probabilities, $o_{d} = 1/\tilde{p}$, of pinnacle sports change at the beggining and end before the match.  

```{r plot_logprobas}
# plot first and last
df_updateodds.pinnacle %>% 
  ggplot(aes(log(1/first_odd), log(1/last_odd))) +
  geom_point(alpha = 0.1) +
  facet_wrap(. ~ type) +
  ggtitle("Dispersión de los momios antes y después", subtitle = "Momios corregidos de Pinnacle Sports") +
  theme_bw() 

```

Look variability by deciles. It grows when there is a clear underdog

```{r look_psdeciles}
# get deciles
cat("deciles for odds home\n")
df_change_odds$psh %>% quantile(1:10/10) 

cat("\ndeciles for odds draw\n")
df_change_odds$psd %>% quantile(1:10/10)

cat("\ndeciles for odds away\n")
df_change_odds$psa %>% quantile(1:10/10)


```

look dispersion by decile of each odd
```{r look_varodds}
# trend
df_varodds %>% 
  ggplot(aes(factor(psh_dscrt_level), variance, shape = type, group = type)) +
  geom_point(size = 8) +
  geom_line(size = 2) +
  xlab("Decil") + ylab("Varianza") +
  ggtitle("Varianza a partir del decil del momio", subtitle = "Diferencia entre el momio inicial y final") +
  theme_minimal() +
  scale_shape_discrete(labels = c("Visitante", "Empate", "Casa")) +
  guides(shape = guide_legend(title = "Tipo de Momio")) +
  theme_bw() 
  
gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 25),
    axis.text.y = element_text(size = 25),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25), 
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 25)
  ) 

# save
ggsave('varianza_momio.png', plot = gph, device = png(), path = "Figures/Football_UK", width = 17.5, height = 10)

```



# Track Take

## Comision of bookmarks

Create tables

```{r df_tracktake}
# create table
df_tt.summary <- df_tt %>% 
  group_by(
    season, books
  ) %>% 
  summarise(
    mean_tt = mean(track_take, na.rm = TRUE),
    stdev_tt = sd(track_take, na.rm = TRUE),
    total_obs = sum(!is.na(track_take))
  ) %>% 
  ungroup(
    
  )

# look tables
df_tt %>% head()
df_tt.summary %>% head()
```

Look comisions of each bookmark

```{r look_comision}
# plot 
df_tt.summary %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE)
  ) %>% 
  ggplot(aes(x = season, y = mean_tt, shape = books, group = books)) +
  geom_point(size = 2) +
  geom_line(linetype = 2) + 
  geom_errorbar(
    aes(
      ymin = mean_tt - qt(p = 1 - 0.01/2, df = total_obs - 1) * stdev_tt / sqrt(total_obs),
      ymax = mean_tt + qt(p = 1 - 0.01/2, df = total_obs - 1) * stdev_tt / sqrt(total_obs)
    ),
    width = 0.2
  ) + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.085), breaks = 0:9/100) + 
  ggtitle("Comisión de las casas de apuestas para la EPL", subtitle = "Intervalo de condianza al 99% para la media de comisión") +
  xlab("Temporada") + ylab("Comisión") +
  guides(shape = guide_legend(title = "Casa de Apuesta")) +
  theme_bw() 

gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 25)
  )

# save
ggsave('comision_casas.png', plot = gph, device = png(), path = "Figures/Football_UK", width = 17.5, height = 10)




```

## Spread of Comisions

Look market efficiency of odds by looking the median of the difference between the max and minimum odd of bookmakers of any type of odd in a game

```{r look_spreads}
df_odds %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE)
  ) %>% 
  ggplot(aes(season, best_spread)) +
  geom_violin(aes(fill = season), alpha = 0.5) +
  geom_boxplot(width = 0.2 ) +
  scale_y_continuous(trans = "log2", breaks = c(0.1, 0.2, 0.5, 2^(0:5))) +
  ggtitle("Spread de momios de la EPL de 6 casas de apuestas", subtitle = "Escala logarítmica base 2") +
  scale_fill_grey() +
  xlab("Temporada") + ylab("Spread") +
  theme_bw() +
  theme(
    legend.position = "none"
  )


gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 25),
    axis.text.y = element_text(size = 25),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    legend.position = "none"
  )

# save
ggsave('spread_momios.png', plot = gph, device = png(), path = "Figures/Football_UK", width = 17.5, height = 10)



```



## Positive track take

```{r tables_tracktake_negative}
# get odds and summaries
df_tt_odds <- df_odds %>% 
  select(
    match_name, season, best_spread, market_tracktake
  ) %>% 
  left_join(
    df_r %>% 
      mutate(
        match_name = paste(hometeam, awayteam, sep = "-")
      ) %>% 
      select(
        match_name, season, psch:psca
      ),
    by = c("match_name", "season")
  ) %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE),
    
    final_tt = 1/psch + 1/pscd + 1/psca - 1,
    
    psch = psch * (final_tt + 1),
    prob = 1/psch
  ) 

df_tt_odds.summary <- df_tt_odds %>% 
  group_by(season) %>% 
  summarise(
    perc_match_goodtt = if_else(market_tracktake < 0, 1, 0) %>% mean(na.rm = TRUE),
    num_match_goodtt = if_else(market_tracktake < 0, 1, 0) %>% sum(na.rm = TRUE),
    num_obs = sum(!is.na(market_tracktake))
  )


# bar plot
df_tt_odds.summary %>% 
  ggplot(aes(season, perc_match_goodtt)) +
  geom_bar(stat = "identity", color = "black", size = 1, fill = "grey65") +
  geom_line(aes(group = 2), size = 2) + 
  geom_label(aes(label = num_match_goodtt), size = 8) +
  scale_y_continuous(labels = scales::percent) + 
  xlab("Temporada") + ylab("Porcentaje") +
  ggtitle("Partidos con comisión negativa", subtitle = "Número de partidos favorables de los 380 posibles") +
  theme_bw() 

gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 25),
    axis.text.y = element_text(size = 25),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    legend.position = "none"
  )

# save
ggsave('arbitraje_tracktake.png', plot = gph, device = png(), path = "Figures/Football_UK", width = 17.5, height = 10)

  
# time series
df_tt_odds.summary %>% 
  ggplot(aes(season, perc_match_goodtt, group = 1)) +
  geom_point(aes(color = season), size = 20) +
  geom_line(aes(color = season), size = 1) +
  geom_label(aes(label = num_match_goodtt), size = 8) +
  scale_y_continuous(labels = scales::percent) + 
  scale_color_grey() + 
  xlab("Temporada") + ylab("Porcentaje") +
  ggtitle("Partidos con arbitraje posible de la EPL", subtitle = "Número de partidos favorables de los 380 posibles") +
  theme_bw() +
  theme(legend.position = "none")



gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 25),
    axis.text.y = element_text(size = 25),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    legend.position = "none"
  )

# save
ggsave('arbitraje_tracktake_timeline.png', plot = gph, device = png(), path = "Figures/Football_UK", width = 17.5, height = 10)

  


```

## Track take and probabilities

Regression between normalized probabilities of pinnacle sports and comision charged

```{r look_tt}
# get R^2 from cuadratic regression
equations <- df_r %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE),
    
    final_tt = 1/psch + 1/pscd + 1/psca - 1,
    
    psch = psch * (final_tt + 1),
    prob = 1/psch
  ) %>% 
  plyr::ddply("season", lm_eqn)

# plot
df_r %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE),
    
    final_tt = 1/psch + 1/pscd + 1/psca - 1,
    
    psch = psch * (final_tt + 1),
    prob = 1/psch
  ) %>% 
  ggplot(aes(prob, final_tt)) +
  geom_point(shape = 1, alpha = 0.5) +
  geom_smooth(formula = y ~ x + I(x^2), method = "lm", color = "black", size = 2, se = FALSE) +
  geom_text(data = equations, aes(x = 0.15, y = 0.05, label = V1), parse = TRUE, inherit.aes = FALSE, size = 6) +
  facet_wrap(season ~ .) + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) + 
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), limits = c(0,1)) +
  ggtitle("Relación entre la comisión y probabilidad", subtitle = "Momios como Local de Pinnacle Sports") +
  xlab("Probabilidad") + ylab("Comisión") +
  theme_bw()

gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 25),
    strip.text.x = element_text(size = 20),
    strip.text.y = element_text(size = 10)
  )

# save
ggsave('regresion_tt.png', plot = gph, device = png(), path = "Figures/Football_UK", width = 17.5, height = 10)

```







