---
title: "EDA SoFIFA"
output: html_notebook
  html_document:
    df_print: paged
---

# Notes

By the fact that the data isn't complete per year, problem approximated in `source\wrangling\tidy_sofifa.R`, the data analysis would be over the data and the approximation made for the missing matchweeks per season. We will join all the final tables from football_uk and understats by their date, in order to make data analysis over the same dates.


# Setup

```{r setup}
library(tidyverse)
library(tidyr)

getwd()
```


# Read data

Read files made in `source\wrangling\`

```{r read_data, warning=FALSE}
# read data
df_understats.r <- read_csv("Data/Understat/final_dbb/understat_finaldbb.csv")
df_football.r <- read_csv("Data/football_uk/final_dbb/footballuk_finaldbb.csv")
df_sofifa.r <- read_csv("Data/SoFIFA/final_dbb/sofifa_finaldbb.csv")

```


# Merge 

First get the distinct names for the same team per table

```{r table_names}
df_teamnames <- tibble(
  team_understats = df_understats.r$team_name %>% unique() %>% sort(),
  team_football = df_football.r$hometeam %>% unique()  %>% sort(),
  team_sofifa = df_sofifa.r$name_team %>% unique()  %>% sort()
) %>% 
  mutate(
    official_name = team_football %>% str_to_lower() %>% str_replace(pattern = " ", replacement = "_"),
    official_name = if_else(official_name == "qpr", "queens_rangers", official_name)
  ) 

df_teamnames

```

Set the same names

```{r set_names}
# for understats
df_understats.r <- df_understats.r %>% 
  left_join(
    df_teamnames %>% select(team_understats, official_name),
    by = c("team_name" = "team_understats")
  )

# for football_uk
df_football.r <- df_football.r %>% 
  left_join(
    df_teamnames %>% select(team_football, official_name),
    by = c("hometeam" = "team_football")
  ) %>% 
  left_join(
    df_teamnames %>% select(team_football, official_name),
    by = c("awayteam" = "team_football"),
    suffix = c(".home", ".away")
  )

# for sofifa
df_sofifa.r <- df_sofifa.r %>% 
  left_join(
    df_teamnames %>% select(team_sofifa, official_name),
    by = c("name_team" = "team_sofifa")
  )

```


# Data wrangle

Lead observations. It doesnt make sense to look how is a team in a given date looking how they did in the same date. So, we should lag observations by a matchweek.

```{r table_lags}
# understats
df_understats <- df_understats.r %>% 
  arrange(team_name, date) %>% 
  group_by(team_name) %>% 
  mutate(
    npxGD_ma = lag(npxGD_ma, n = 1, default = NA),
    npxGD_var = lag(npxGD_var, n = 1, default = NA),
    year_date = lubridate::year(date) - 2000, # creating keys for future join between tables
    week_date = lubridate::isoweek(date),
    season = year - 2000
  ) %>% 
  ungroup() %>% 
  select(
    season, matchweek, official_name,
    position_table:npxGD_var
  ) %>% 
  rename(
    team = official_name
  )

# football_uk
df_football <- df_football.r %>% 
  group_by(season) %>% 
  mutate(
    matchweek = ((row_number() -1) %/% 10) + 1 # create matchweek (are sorted)
  ) %>% 
  ungroup() %>% 
  select(
    season, date, matchweek,
    official_name.home, official_name.away, result,
    starts_with("max_odds"), market_tracktake, psch:psca,
    starts_with("diff")
  ) %>% 
  rename(
    hometeam = official_name.home,
    awayteam = official_name.away
  )

# sofifa
df_sofifa <- df_sofifa.r %>% 
  mutate(
    season = str_extract(fifa, "[[:digit:]]+") %>% as.integer() - 1
  ) %>% 
  select(
    season, week_date, official_name, ova:was_na
  ) %>%
  rename(
    team = official_name
  )

```


Join data in a main dataframe

```{r get_dfmain}
df_main <- df_football %>% 
  inner_join(
    df_understats,
    by = c("season", "matchweek", "hometeam" = "team")
  ) %>% 
  inner_join(
    df_understats ,
    by = c("season", "matchweek", "awayteam" = "team"),
    suffix = c(".home", ".away")
  ) %>% 
  mutate(
    week_date = lubridate::isoweek(date)
  ) %>% 
  inner_join(
    df_sofifa,
    by = c("season", "week_date", "hometeam" = "team")
  ) %>% 
  inner_join(
    df_sofifa %>% select(-was_na),
    by = c("season", "week_date", "awayteam" = "team"),
    suffix = c(".home", ".away")
  ) 

df_main %>% head()
```


# Exploratory Data Analysis

## Data sanity

There is going to be 1 less matchweek by the fact of the `NA` induced by `lag(npxGD_*)` in the `inner_join()`.

Look the number of `NA` in sofifa weeks. All ok.

```{r look_na}
df_main %>% 
  group_by(season) %>% 
  summarise(
    prop_na = sum(was_na) / n()
  )

```


## Stats

Get the mean and standard dev. for the grades of the teams. First get aggregates

```{r get_summarystats}
# how much changes sofifa variables between teams per season #
# get data
df_weeks <- df_main %>% 
  select(
    season, week_date
  ) %>% 
  unique()

# get with all the matches
df_sofifa.used_data <- df_sofifa %>% 
  semi_join(
    df_weeks,
    by = c("season", "week_date")
  )

# aggregate data
df_stats_sofifa <- df_sofifa.used_data %>% 
  group_by(
    season, team
  ) %>% 
  summarise_at(
    vars(ova:taa),
    list(mean=mean, sd=sd),
    na.rm = TRUE
  ) %>% 
  left_join(
    df_weeks %>% count(season) %>% rename(num_obs = n), # 
    by = "season"
  ) %>% 
  ungroup(
    
  )

# pivot longer
df_stats_sofifa.mean <-  df_stats_sofifa %>% 
  select(
    season, team, ova_mean:def_mean
  ) %>% 
  pivot_longer(
    cols = ova_mean:def_mean,
    names_to = "type_stat",
    values_to = "mean_value"
  ) %>% 
  mutate(
    type_stat = str_replace(type_stat, pattern = "_mean", replacement = "")
  )

df_stats_sofifa %>% head()
```

Next, get mean and std

```{r get_mean_std}
df_stats_sofifa.mean <-  df_stats_sofifa %>% 
  select(
    season, team, ova_mean:def_mean
  ) %>% 
  pivot_longer(
    cols = ova_mean:def_mean,
    names_to = "type_stat",
    values_to = "mean_value"
  ) %>% 
  mutate(
    type_stat = str_replace(type_stat, pattern = "_mean", replacement = "")
  )

df_stats_sofifa.sd <-  df_stats_sofifa %>% 
  select(
    season, team, ova_sd:def_sd, num_obs
  ) %>% 
  pivot_longer(
    cols = ova_sd:def_sd,
    names_to = "type_stat",
    values_to = "std_value"
  ) %>% 
  mutate(
    type_stat = str_replace(type_stat, pattern = "_sd", replacement = "")
  )

df_stats_sofifa.sd
df_stats_sofifa.mean

```



Plot mean with sd of OVA, ATT, MEAN, DEF

```{r plt_teams}
df_stats_sofifa.mean %>% 
  left_join(
    df_stats_sofifa.sd,
    by = c("season", "team", "type_stat")
  ) %>% 
  filter(
    team %in% c("arsenal", "chelsea",  "liverpool", "man_united", "man_city", "tottenham")
  ) %>% 
  mutate(
    team = case_when(
      team == "man_united" ~ "United",
      team == "man_city" ~ "City",
      TRUE ~ team
    ),
    team = str_to_title(team),
    type_stat = case_when(
      type_stat == "att" ~ "Ataque",
      type_stat == "mid" ~ "Media",
      type_stat == "def" ~ "Defensa",
      type_stat == "ova" ~ "General"
    ),
    type_stat = factor(type_stat, levels = c("General", "Ataque", "Media", "Defensa"), ordered = TRUE)
  ) %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE)
  ) %>% 
  ggplot(aes(x = season, y = mean_value, shape = type_stat, group = type_stat)) +
  geom_point(size = 2) +
  geom_line(aes(linetype = type_stat)) + 
  geom_errorbar(
    aes(
      ymin = mean_value - qt(p = 1 - 0.01/2, df = num_obs - 1) * std_value / sqrt(num_obs),
      ymax = mean_value + qt(p = 1 - 0.01/2, df = num_obs - 1) * std_value / sqrt(num_obs)
    ),
    width = 0.2
  ) + 
  scale_linetype_manual(values = c("solid", "dashed", "dashed", "dashed")) +
  scale_shape_manual(values = c(19, 0, 1, 2)) +
  ggtitle("Calificación de EA Sports de los equipos de la EPL", subtitle = "Intervalo de confianza al 99% para la media de calificaciones") +
  xlab("Temporada") + ylab("Calificación") +
  guides(shape = guide_legend(title = "Tipo"), linetype = guide_legend(title = "Tipo")) +
  theme_bw() +
  facet_wrap(team ~ .)

gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 25),
    strip.text.x = element_text(size = 20)
  ) 


# save
ggsave('general_stats_sofifa.png', plot = gph, device = png(), path = "Figures/SoFIFA/", width = 17.5, height = 10)

```


Tranfer market

```{r get_transfer}
# value of team per season
df_means_transfer <- df_stats_sofifa %>% 
  group_by(season) %>% 
  summarise(mean_transfer = mean(transfer_budget_mean))

df_stats_sofifa.transfer <- df_stats_sofifa %>% 
  add_row(
    season = df_means_transfer$season,
    transfer_budget_mean = df_means_transfer$mean_transfer,
    team = "mean_team"
  ) %>% 
  mutate(
    type_data = ifelse(team == "mean_team", 1, 0) 
  )

df_stats_sofifa.transfer %>% head()

```


```{r plt_transfer}
df_stats_sofifa.transfer %>% 
  mutate(
    season = factor(season, levels = 13:20, ordered = TRUE)
  ) %>% 
  ggplot(aes(season, transfer_budget_mean, group = team)) +
  geom_point(aes(shape = factor(type_data)), size = 4) +
  geom_line(aes(linetype = factor(type_data))) +
  ggrepel::geom_text_repel(
    data = df_stats_sofifa.transfer %>% filter(season == 20) %>%  mutate(season = factor(season, levels = 13:20, ordered = TRUE)),
    aes(label = team),
    nudge_x = 10,
    direction = "y",
    size = 5
  ) +
  ggrepel::geom_label_repel(
    data = df_stats_sofifa.transfer %>% filter(team == "mean_team") %>%  mutate(season = factor(season, levels = 13:20, ordered = TRUE)),
    aes(label = scales::dollar(transfer_budget_mean, accuracy = 2, suffix = "€", prefix = "")),
    direction = "y",
    nudge_y = 5,
    size = 5
  ) +
  guides(size = FALSE, shape = guide_legend("Tipo"), linetype = guide_legend("Tipo")) +
  scale_shape_manual(values = c(1, 19), labels = c("Equipo", "Promedio")) +
  scale_linetype_manual(values = c("dashed", "solid"), labels = c("Equipo", "Promedio")) +
  scale_y_continuous(labels = scales::dollar_format(accuracy = 2, suffix = "€", prefix = "")) +
  xlab("Temporada") + ylab("Millones de euros") +
  ggtitle("Presupuesto del equipo en el mercado de transferencias") +
  theme_bw() 

gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 25),
    legend.position = "bottom",
    strip.text.x = element_text(size = 20)
  ) 

# save
ggsave('transfer_budget.png', plot = gph, device = png(), path = "Figures/SoFIFA/", width = 17.5, height = 10)

  
```




























