---
title: "EDA Understats"
output: html_notebook
  html_document:
    df_print: paged
---

# Setup

```{r setup}
library(tidyverse)
library(tidyr)

# knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_knit$set(root.dir = '') 
getwd()
```

# Read data

Read files made in `source\wrangling\`

```{r read_data, warning=FALSE}
# folder
folder_files <- "Data/Understat/final_dbb/"

# read tables
df_r <- read_csv(file = paste0(folder_files, "understats_raw.csv"))
df_means <- read_csv(file = paste0(folder_files, "means_xi.csv"))
df_final <- read_csv(file = paste0(folder_files, "understat_finaldbb.csv"))

# final values
# colors
COLS_BIGSIX <- c("#9c824a", "#034694", "#00a398", "#97c1e7", "#da020e", "#132257")

```


# General stats

## Base rates

EPL final season table
```{r tbl_epl_seasons}
# table 
df_r %>%
  group_by(year, team_name) %>%
  transmute(
    # select important vars
    year, matchweek, team_name, 

    # select important vars to compute aggregates
    scored, missed, pts,
    
    # running statistics by season 
    total_pts = cumsum(pts),
    diff_goals = cumsum(scored - missed),
    total_goals = cumsum(scored)
    
  ) %>% 
  ungroup() %>% 
  arrange(
    # premier league orders by: points, goal difference, total goals
    year, matchweek, desc(total_pts), desc(diff_goals), desc(total_goals)
  ) %>% 
  group_by(year, matchweek) %>% 
  mutate(
    position_table = dense_rank(
        desc(interaction(total_pts, diff_goals, total_goals, lex.order = TRUE))
      )
  ) %>% 
  filter(
    matchweek == 38
  ) 


```

Win-draw-loss per seasons by local and away team and general
```{r}
df_r %>% 
  group_by(year, h_a) %>% 
  summarise(
    home = mean(wins),
    draw = mean(draws),
    away = mean(loses)
  ) %>% 
  filter(h_a == "h") %>% 
  select(-h_a) 

# latex code of last table
# df_r %>% 
#   group_by(year, h_a) %>% 
#   summarise(
#     home = mean(wins),
#     draw = mean(draws),
#     away = mean(loses)
#   ) %>% 
#   filter(h_a == "h") %>% 
#   select(-h_a) %>% 
#   xtable::xtable(type = "latex")

df_r %>% 
  group_by(h_a) %>% 
  summarise(
    home = mean(wins),
    draw = mean(draws),
    away = mean(loses)
  )
```


Win-draw-loss rates by team
```{r tbl_wdl_teams}
df_r %>% 
  group_by(team_name, h_a) %>% 
  summarise(
    win = mean(wins),
    draw = mean(draws),
    loss = mean(loses)
  ) %>% 
  pivot_wider(
    names_from = "h_a",
    values_from = win:loss
  )

```

All epl teams from season 2014 to season 2020 

```{r}
df_r$team_name %>% unique()
```


# npXG

Non-penalty expected goals along time. LOESS fit.

```{r npxg_smooth}
df_r %>% 
  filter(
    team_name %in% c("Manchester United", "Manchester City", "Liverpool",
                     "Arsenal", "Chelsea", "Tottenham")
  ) %>% 
  ggplot(aes(matchweek, npxG, color = team_name, group = year)) +
  geom_smooth(se = TRUE, span = 0.5) +
  geom_point(alpha = 0.5) +
  facet_grid(team_name ~ year) +
  scale_color_manual(values = COLS_BIGSIX) +
  ggtitle("Goles esperados (sin penales) del Big Six", subtitle = "Temporada 2014-2020") +
  xlab("Jornada") + ylab("npxG") + 
  theme_bw() +
  theme(
    legend.position  = "none"
  )

```

# Diff npXG

LOESS non penalty expected goals difference between them and the rival

```{r}
df_r %>% 
  filter(
    team_name %in% c("Manchester United", "Manchester City", "Liverpool",
                     "Arsenal", "Chelsea", "Tottenham")
  ) %>% 
  ggplot(aes(matchweek, npxG - npxGA, color = team_name, group = year)) +
  geom_smooth(se = TRUE, span = 0.5) +
  geom_point(alpha = 0.5) +
  facet_grid(team_name ~ year) +
  ggtitle("Diferencia de goles esperados del Big Six", subtitle = "xG sin penales") +
  geom_hline(yintercept = 0, linetype = 2) +
  xlab("Jornada") + ylab("Diferencia npxG") + 
  scale_color_manual(values = COLS_BIGSIX) +
  theme_bw() +
  theme(
    legend.position  = "none" 
  )

```

Density per season of each big six

```{r npxg_dsty}
df_r %>% 
  filter(
    team_name %in% c("Manchester United", "Manchester City", "Liverpool",
                     "Arsenal", "Chelsea", "Tottenham")
  ) %>% 
  mutate(
    team_name = case_when(
      team_name == "Manchester United" ~ "United",
      team_name == "Manchester City" ~ "City",
      TRUE ~ team_name
    ),
    year = year - 2000
  ) %>% 
  ggplot(aes(npxG - npxGA, group = year)) +
  geom_density(fill = "grey60") + 
  facet_grid(team_name ~ year) +
  ggtitle("Diferencia de goles esperados del Big Six", subtitle = "xG sin penales") +
  geom_vline(xintercept = 0, linetype = 2) +
  xlab("Diferencia npxG") + ylab("") + 
  theme_bw() +
  theme(
    legend.position  = "none"
  )

gph <- last_plot() +
    theme(
    axis.text.x = element_text(size = 15),
    axis.text.y = element_blank(),
    axis.title.x = element_text(size = 25),
    title = element_text(size = 25),
    strip.text.x = element_text(size = 20),
    strip.text.y = element_text(size = 10),
    legend.position  = "none"
  )


# save
ggsave('npxGD_density.png', plot = gph, device = png(), path = "Figures/Understats/", width = 17.5, height = 10)

```

Boxplot of the npXGD of the Big Six

```{r npxg_boxplot}
df_r %>% 
  filter(
    team_name %in% c("Manchester United", "Manchester City", "Liverpool",
                     "Arsenal", "Chelsea", "Tottenham")
  ) %>% 
  ggplot(aes(factor(year), npxG - npxGA, color = team_name)) +
  geom_boxplot(position = "dodge2", size = 2) + 
  geom_hline(yintercept = 0, linetype = 2) +
  scale_color_manual(values = COLS_BIGSIX) +
  ggtitle("Diferencia de goles esperados del Big Six", subtitle = "xG sin penales") +
  xlab("Temporada") + ylab("Diferencia npxG") + 
  theme_bw() +
  guides(color = guide_legend(title = "Equipo")) +
  theme(
    legend.position = "bottom",
  )
```

# Ponderated stats

decayment of the $\lambda_{j}$ weights


```{r weight_decayment}
expand.grid(
  t = 1:100,
  xi = c(0, 0.01, 0.05, 0.1, 0.25, 0.5, 1, 10, Inf) 
) %>% 
  mutate(
    decay = exp(- xi * t),
    xi = factor(xi, ordered = TRUE)
  ) %>% 
  ggplot(aes(t, decay, color = xi, group = xi)) +
  geom_line(size = 1.5) +
  scale_y_continuous() +
  xlab("Tiempo") + ylab("Peso") +
  ggtitle("Decaimiento exponencial de las observaciones", subtitle = "Por fuerza de decaimiento") +
  theme_bw()


```

Ponderated mean $\bar{X}_{\lambda} = \sum_{j} \lamda_{j} X_{j}$

```{r}
df_means %>% 
  filter(
    team_name %in% c("Aston Villa", "Leicester", "Manchester City","Tottenham")
  ) %>% 
  mutate(
    xi = str_replace(xi, pattern = "wmean_", replacement = ""),
    xi = factor(xi, labels = c("\u03be = 0", "\u03be = 0.05", "\u03be = 0.1", "\u03be = 0.5",  "\u03be = \u221e"), ordered = TRUE)
  ) %>% 
  ggplot(aes(date, mean_value, group = xi)) +
  geom_point(
    aes(y=npxGD),
    color = "gray70",
    shape = 1
  ) +
  geom_hline(yintercept = 0, linetype = 2, color = "gray70") +
  geom_line() +
  facet_grid(xi ~ team_name) +
  scale_y_continuous(limits = c()) +
  scale_x_date(
    breaks = seq(as.Date("2014-01-01"), as.Date("2021-01-01"), by = "years"),
    date_labels = "%Y"
  ) +
  ggtitle("Media ponderada para los npxGD de la EPL", subtitle = "Variando el valor de \u03be") +
  xlab("Fecha") + ylab(latex2exp::TeX("$\\bar{x}_{\\lambda}")) +
  theme_bw() 


gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 15, angle = 45, vjust = 0.5),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    strip.text.x = element_text(size = 20),
    strip.text.y = element_text(size = 20)
  )

# save
ggsave('npxGD_meanXi.png', plot = gph, device = png(), path = "Figures/Understats/", width = 17.5, height = 10)


```

Ponderated variance $s^{2}_{\lambda} = \sum_{j}\frac{\lambda_{j}}{1 - \sum_{k}\lambda_{k}^{2}}(X_{j} - \bar{X}_{\lambda})^{2}=$ of the big six with $\xi = 0.01$.

```{r npxgd_meanvar}
# get df at begin and end
df_beginend <- df_r %>% 
  group_by(year) %>% 
  summarise(
    start_season = min(date),
    end_season = max(date)
  )


# plot
df_final %>% 
  filter(
    team_name %in% c("Aston Villa", "Leicester", "Manchester City", "Tottenham")
  ) %>%
  ggplot(aes(x = date, y = npxGD_ma)) +
  geom_ribbon(
    aes(
      ymin = npxGD_ma - (sqrt(npxGD_var)),
      ymax = npxGD_ma + (sqrt(npxGD_var))
    ),
    fill = "gray70"
  ) +
  geom_line(aes(group = 1), color = "black", size = 1) +
  geom_vline(xintercept = df_beginend$start_season, linetype = 2, alpha = 0.5) +
  geom_vline(xintercept = df_beginend$end_season, linetype = 1, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 1, alpha = 0.5) +
  geom_label(aes(x = as.Date("2014-08-16"), y = 4.7, label = c("inicio")), size = 6) +
  geom_label(aes(x = as.Date("2015-05-24"), y = 4.7, label = c("fin")), size = 6) +
  ggrepel::geom_text_repel(
    data = tibble(
      date = as.Date("2020-05-01"),
      team_name = c("Aston Villa", "Leicester", "Manchester City", "Tottenham"),
      npxGD_ma = c(-0.8, 0.30, 1.25, -0.41)
    ),
    aes(label = "Covid-19"),
    arrow = arrow(length = unit(0.015, "npc")),
    point.padding = 0.2,
    nudge_x = .15,
    nudge_y = 1.5,
    size = 6,
    segment.size = 1
  ) +
  facet_wrap( ~ team_name) +
  scale_x_date(
    minor_breaks = "2 months", 
    breaks = c(df_beginend$end_season, df_beginend$start_season),
    date_labels = "%Y-%m"
  ) +
  ggtitle("Serie de tiempo de la media ponderada del npxGD", subtitle = "Más/menos una desviación estándar ponderada") +
  xlab("Fecha") + ylab("Media") + 
  theme_bw() 

gph <- last_plot() +
  theme(
    axis.text.x = element_text(size = 15, angle = 45, vjust = 0.5),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(size = 25),
    axis.title.y = element_text(size = 25),
    title = element_text(size = 25),
    strip.text = element_text(size = 17)
  )

# save
ggsave('npxGD_meanvar.png', plot = gph, device = png(), path = "Figures/Understats/", width = 17.5, height = 10)



```


